---
title: 'Inference for numerical data'
output:
  html_document:
    css: ./lab.css
    highlight: pygments
    theme: cerulean
  pdf_document: default
---

## North Carolina births

In 2004, the state of North Carolina released a large data set containing 
information on births recorded in this state. This data set is useful to 
researchers studying the relation between habits and practices of expectant 
mothers and the birth of their children. We will work with a random sample of 
observations from this data set.

## Exploratory analysis

Load the `nc` data set into our workspace.

```{r load-data, eval=TRUE}
load("more/nc.RData")
head(nc)
```

We have observations on 13 different variables, some categorical and some 
numerical. The meaning of each variable is as follows.

variable         | description
---------------- | -----------
`fage`           | father's age in years.
`mage`           | mother's age in years.
`mature`         | maturity status of mother.
`weeks`          | length of pregnancy in weeks.
`premie`         | whether the birth was classified as premature (premie) or full-term.
`visits`         | number of hospital visits during pregnancy.
`marital`        | whether mother is `married` or `not married` at birth.
`gained`         | weight gained by mother during pregnancy in pounds.
`weight`         | weight of the baby at birth in pounds.
`lowbirthweight` | whether baby was classified as low birthweight (`low`) or not (`not low`).
`gender`         | gender of the baby, `female` or `male`.
`habit`          | status of the mother as a `nonsmoker` or a `smoker`.
`whitemom`       | whether mom is `white` or `not white`.

1.  What are the cases in this data set? How many cases are there in our sample?

As a first step in the analysis, we should consider summaries of the data. This
can be done using the `summary` command:

```{r summary, eval=FALSE}
summary(nc)
nrow(nc)  #this gives us the number of rows as well as the number of cases
```

As you review the variable summaries, consider which variables are categorical 
and which are numerical. For numerical variables, are there outliers? If you 
aren't sure or want to take a closer look at the data, make a graph.

```{r plotting}

boxplot(nc$fage)
boxplot(nc$mage)
boxplot(nc$weeks)
boxplot(nc$visits)
boxplot(nc$gained)
boxplot(nc$weight)

#looking at the boxplots for the numerical categories, it looks as though they all have some outliers
```

Consider the possible relationship between a mother's smoking habit and the 
weight of her baby. Plotting the data is a useful first step because it helps 
us quickly visualize trends, identify strong associations, and develop research
questions.

2.  Make a side-by-side boxplot of `habit` and `weight`. What does the plot 
highlight about the relationship between these two variables?

```{r boxplot}
boxplot(nc$weight ~ nc$habit)

#the boxplot shows that the median is about the same but non smokers have wider range.
```

The box plots show how the medians of the two distributions compare, but we can
also compare the means of the distributions using the following function to 
split the `weight` variable into the `habit` groups, then take the mean of each
using the `mean` function.

```{r by-means, eval=FALSE}
by(nc$weight, nc$habit, mean)
```

There is an observed difference, but is this difference statistically 
significant? In order to answer this question we will conduct a hypothesis test
.

## Inference

3.  Check if the conditions necessary for inference are satisfied. Note that 
you will need to obtain sample sizes to check the conditions. You can compute 
the group size using the same `by` command above but replacing `mean` with 
`length`.


```{r conditions}
by(nc$weight, nc$habit, length)
```

* The conditions for inference are independence/sample size and normality. Our sample size is from less than 10% of population. Because of our large sample size, we can relax the nearly normal condition.

4.  Write the hypotheses for testing if the average weights of babies born to 
smoking and non-smoking mothers are different.

* NULL: ($\mu_{nonsmoker} - \mu_{smoker}$) = 0
* ALTERNATE: ($\mu_{nonsmoker} - \mu_{smoker}$) != 0


Next, we introduce a new function, `inference`, that we will use for conducting
hypothesis tests and constructing confidence intervals. 

```{r inf-weight-habit-ht, eval=FALSE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ht", null = 0, 
          alternative = "twosided", method = "theoretical")
```

Let's pause for a moment to go through the arguments of this custom function. 
The first argument is `y`, which is the response variable that we are 
interested in: `nc$weight`. The second argument is the explanatory variable, 
`x`, which is the variable that splits the data into two groups, smokers and 
non-smokers: `nc$habit`. The third argument, `est`, is the parameter we're 
interested in: `"mean"` (other options are `"median"`, or `"proportion"`.) Next
we decide on the `type` of inference we want: a hypothesis test (`"ht"`) or a 
confidence interval (`"ci"`). When performing a hypothesis test, we also need 
to supply the `null` value, which in this case is `0`, since the null 
hypothesis sets the two population means equal to each other. The `alternative`
hypothesis can be `"less"`, `"greater"`, or `"twosided"`. Lastly, the `method` 
of inference can be `"theoretical"` or `"simulation"` based.

5.  Change the `type` argument to `"ci"` to construct and record a confidence 
interval for the difference between the weights of babies born to smoking and 
non-smoking mothers.

```{r  ci}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical")

#95% confidence interval is: ( 0.0534 , 0.5777 )
```


By default the function reports an interval for ($\mu_{nonsmoker} - \mu_{smoker}$)
. We can easily change this order by using the `order` argument:

```{r inf-weight-habit-ci, eval=FALSE, tidy=FALSE}
inference(y = nc$weight, x = nc$habit, est = "mean", type = "ci", null = 0, 
          alternative = "twosided", method = "theoretical", 
          order = c("smoker","nonsmoker"))
```

* * *

## On your own

-   Calculate a 95% confidence interval for the average length of pregnancies 
(`weeks`) and interpret it in context. Note that since you're doing inference 
on a single population parameter, there is no explanatory variable, so you can 
omit the `x` variable from the function.

```{r ci weeks}
inference(nc$weeks, est = "mean", type = "ci", null = 0, alternative = "twosided", method = "theoretical")

#95% CI = (38.1528, 38.5165)
```

-   Calculate a new confidence interval for the same parameter at the 90% 
confidence level. You can change the confidence level by adding a new argument 
to the function: `conflevel = 0.90`.

```{r ci 90}
inference(nc$weeks, est = "mean", type = "ci", null = 0, alternative = "twosided", method = "theoretical", conflevel = 0.90 )

#90% CI: (38.182, 38.4873)
```

-   Conduct a hypothesis test evaluating whether the average weight gained by 
younger mothers is different than the average weight gained by mature mothers.

```{r Hypotheses}
inference(y = nc$weight, x = nc$mature, est = "mean", type = "ht", null = 0, alternative = "twosided", method = "theoretical" ) 

# our p-value, .8526 > .05 so we fail to reject the null hypothesis which was that there was no difference in the weights 
```

-   Now, a non-inference task: Determine the age cutoff for younger and mature 
mothers. Use a method of your choice, and explain how your method works.

```{r max age}
#create subsets of mature and younger moms and verify that this was done
mature <- subset(nc, mature == "mature mom") #
younger <- subset(nc, mature == "younger mom")
View(mature)
View(younger)
#find the max and min age for all the mature moms
maxmat.age <- max(mature$fage, na.rm = TRUE)
maxmat.age
minmat.age <- min(mature$fage, na.rm = TRUE)
minmat.age
#find the max and min age for all the younger moms
maxyng.age <- max(younger$fage, na.rm = TRUE)
maxyng.age
minyng.age <- min(younger$fage, na.rm = TRUE)
minyng.age

#This showed something quite interesting. I'm not certain what criteria was used to determine mature vs. younger but it does not seem to be  purely age since the range of both overlap.
```


-   Pick a pair of numerical and categorical variables and come up with a 
research question evaluating the relationship between these variables. 
Formulate the question in a way that it can be answered using a hypothesis test
and/or a confidence interval. Answer your question using the `inference` 
function, report the statistical results, and also provide an explanation in 
plain language.

* I would like to explore if whether a baby is considered low birth weight is impacted by how much weight the mother gains during pregnancy. In plain terms, is there is a difference in average weight gained in mother with low birth weight babies and mothers with babies who are not considered low birth weight.

    + NULL: ($\mu_{low} - \mu_{notlow}$) = 0
    + ALT: ($\mu_{low} - \mu_{notlow}$) != 0

```{r mynum_cat}
inference(y = nc$gained, x = nc$lowbirthweight, est = "mean", type = "ht", null = 0, alternative = "twosided", method = "theoretical" ) 

#our p-value, .0014 < .05 so we reject our null hypothesis meaning there is enough evidence in the data to say that there is some difference in the average weight gained by mothers whose babies were considered low birth weight and those not.
```

<div id="license">
This is a product of OpenIntro that is released under a [Creative Commons 
Attribution-ShareAlike 3.0 Unported](http://creativecommons.org/licenses/by-sa/3.0).
This lab was adapted for OpenIntro by Mine &Ccedil;etinkaya-Rundel from a lab 
written by the faculty and TAs of UCLA Statistics.
</div>
